A Road Segment Prioritization Approach for Cycling Infrastructure

### 2022-08-31 README.Rmd
proofreading

The funding on its own is therefore no guarantee of a change in
commuting across the country; it must be used to design adequate cycling
infrastructure that is based on motivators and deterrents to cycling. In this paper we 
propose an approach for planning a connected cycling network that is meant to 
be applicable to any city, and we use the city of Manchester as a case study. Greater 
Manchester aims to become the first region to reach a carbon neutral transport network; 
Its local authorities have developed plans for a 1800 mile walking and cycling network 
(Bee Network) that will be delivered over 10 years [@manchestercyclingbee2020]. While schemes 
have already been put in place for around a quarter of this network, the majority is yet to be built.

@@

*Optimization* methods have been used to propose improvements to
cycling networks. @mesbah2012bilevel propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as "traffic evaporation" [@nello2020environmental].
On the other hand, @mauttone2017bicycle developed an optimization framework aiming at minimizing the total
user cost of cycling on the network without considering car usage. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various Origin-Destination (OD) pairs.] on road segments (links) is obtained by using
shortest paths to route existing cycling demand onto the road network. The solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these approaches is that they do not explicitly solve for
continuity. The latter is addressed using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].

In this paper, continuity is analysed by looking at the connectivity of
the network through the graph-theoretic concept of *connected
components*. @orozco2020data for example,
study the existing cycling network in terms of its disconnected
components. They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor. By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is
also at the core of the methodology proposed by @olmos2020data. After
routing the cycling demand onto the network links, they use percolation
theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for
filtering to identify the minimum flow at which the whole city is
connected by a giant component. While these approaches deal with
continuity better, they look at the network as a whole when attempting
to improve it, and in doing so fail to account for equitable
distribution of infrastructure.

@@

This research attempts to propose an egalitarian framework for cycling
network design. This is done by identifying the different sub-networks
that exist within the larger network, and ensuring that each gets a fair
share of investment. Trip patterns in a city are not uniformly
distributed geographically, and *community finding* methods have been
used to partition study areas into localized areas that experience a
disproportionate number of trips within them. @akbarzadeh2018designing
use a modularity maximization approach [@blondel2008fast] on taxi trip
data to identify 7 different communities in the city of Isfahan, Iran.
An optimization problem is then formulated to connect nodes within each
community with cycling infrastructure, with the emphasis being on
connectivity within the communities, not between them. @bao2017planning
adopt a similar methodology, first identifying communities and then
using a greedy network expansion algorithm to simultaneously add links
to each community. The link with the highest benefit-cost ratio in each
community is selected, and the network is grown by adding neighboring
links to the solution until a budget limit is met. The benefit is the
flow on the link, and each link is assigned a cost based on current road
conditions.

@@

Our work builds on these community finding approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
*within communities*. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose an approach that ensures a better distribution 
investments across the study area while also accounting for motivators and deterrents to cycling. 
We propose three sub-methods that address some of the limitations of previous
studies. These limitations include (a) an inherent bias of basing network
design solely on existing cycling demand, (b) the proposal of routes that may
not correspond to studies on cyclist preference and government policies,
and (c) an insufficient consideration of the ethical principles underlying
the analysis. The current work addresses these limitations through the following organisation. Section \@ref(calculating-potential-cycling-demand)
focuses on calculating potential cycling demand. Section \@ref(routing)
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\@ref(road-segment-prioritization) outlines a method for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
'egalitarianism' to one grounded in 'utilitarianism'.

@@

*OD* data can be obtained from a range of sources,
including GPS data, household travels surveys and Census data on work
locations. In areas where observed OD data is unavailable, modelling
techniques such as spatial interaction models can be used to estimate
travel volumes between zones [@black1995spatial; @martinez2013new;
@wilson1971family]. In this paper we use open access data from the UK
census [@ONS2011flowdata], which contains aggregate statistics on
number of commuters between administrative zones --- Middle layer Super
Output Areas (MSOA) --- by mode of travel. MSOAs have an average
population of just over 8000 [@ofn2018population].^[ See
<https://wicid.ukdataservice.ac.uk/> for open access to the OD data.]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling)
illustrate the proportion of trips cycled by distance and the
geographic extent of the input OD dataset used in this paper.

@@

\noindent where $d$ and $s$ are the distance and slope associated with each
OD pair, and $\alpha$ and all $\beta$s, are parameters calculated by a regression
model on the training data. The square and square-root distance terms
"capture the non-linear impact of distance on the likelihood of
cycling", and interaction terms to capture the combined effect of slope
and distance [@lovelace2017propensity]. Alternative cycling uptake
models could be 'plugged in' to our approach for different contexts or scenarios of change.

@@

For the purposes of this paper, we created a similar weighting profile
that is adjusted to favor less stressful roads (based on the definition
by @sorton1994bicycle and information from Table
\@ref(table:osmroadtypes)), and roads with existing cycling
infrastructure. We believe this to be more appropriate than the simplistic
approach adopted by @mauttone2017bicycle, as it makes use of a hierarchy
of road preference based on perceived stress levels, going beyond simply
favoring roads with existing cycling infrastructure.

@@

A range of algorithms could be used for prioritization. Because policy
priorities vary, we present two algorithms. Both utilize existing
infrastructure from the beginning and allow us to compare a solution
that focuses on utilitarianism to one that focuses on egalitarianism. In
both algorithms, links are selected iteratively and the iteration at
which each link is added to the solution is recorded. Investments in
cycling infrastructure can be limited by budget constraints, so it can
be useful to see where best to allocate a defined length of segregated
infrastructure. In order to incorporate egalitarian principles in our
approach, we use community detection to partition the study area and
evaluate the distribution of investment over the different subdivisions.

@@

Consistent growth can also be seen for the size of the Largest Connected
Component in the proposed bicycle network (Figure
\@ref(fig:componentsandGCC)). Here however, we find that there is little
difference between both algorithms.

>>>>

The funding on its own is therefore no guarantee of a change in
commuting across the country; it must be used to design adequate cycling
infrastructure that is based on motivators and deterrents to cycling. In this paper we 
propose an approach for planning a connected cycling network that is meant to 
be applicable to any city (as shown in the Appendix), and we use the city of Manchester as a case study. Greater 
Manchester aims to become the first region to reach a carbon neutral transport network; 
its local authorities have developed plans for a 1800 mile walking and cycling network 
(Bee Network) that will be delivered over 10 years [@manchestercyclingbee2020]. While schemes 
have already been put in place for around a quarter of this network, the majority is yet to be built.

@@

*Optimization* methods have been used to propose improvements to
cycling networks. @mesbah2012bilevel propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as "traffic evaporation" [@nello2020environmental].
On the other hand, @mauttone2017bicycle developed an optimization framework aiming at minimizing the total
user cost of cycling on the network without considering car usage. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various Origin-Destination (OD) pairs.] on road segments (links) is obtained by using
shortest paths to route existing cycling demand onto the road network. The solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these approaches is that they do not explicitly solve for
continuity. Continuity is addressed using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].

In this paper, continuity is analysed by looking at the connectivity of
the network through the graph-theoretic concept of *connected
components*. @orozco2020data study the existing cycling network in terms of its disconnected
components. They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor. By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is
also at the core of the methodology proposed by @olmos2020data. After
routing the cycling demand onto the network links, they use percolation
theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for
filtering to identify the minimum flow at which the whole city is
connected by a giant component. While these approaches deal with
continuity better, they look at the network as a whole when attempting
to improve it, and in doing so fail to account for equitable
distribution of infrastructure.

@@

This research attempts to propose an egalitarian framework for cycling
network design. This is done by identifying the different sub-networks
that exist within the larger network, and ensuring that each gets a fair
share of investment. Trip patterns in a city are not uniformly
distributed geographically, and *community finding* methods have been
used to partition study areas into localized areas that experience a
disproportionate number of trips within them. @akbarzadeh2018designing
use a modularity maximization approach [@blondel2008fast] on taxi trip
data to identify seven different communities in the city of Isfahan, Iran.
An optimization problem is then formulated to connect nodes within each
community with cycling infrastructure, with the emphasis being on
connectivity within the communities, not between them. @bao2017planning
adopt a similar methodology, first identifying communities and then
using a greedy network expansion algorithm to simultaneously add links
to each community. The link with the highest benefit-cost ratio in each
community is selected, and the network is grown by adding neighboring
links to the solution until a budget limit is met. The benefit is the
flow on the link, and each link is assigned a cost based on current road
conditions.

@@

Our work builds on these community finding approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
*within communities*. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose an approach that ensures a better distribution 
investments across the study area while also accounting for motivators and deterrents to cycling. 
We propose three sub-methods that address some of the limitations of previous
studies. These limitations include (a) an inherent bias when basing network
design solely on existing cycling demand, (b) the proposal of routes that may
not correspond to studies on cyclist preference and government policies,
and (c) an insufficient consideration of the ethical principles underlying
the analysis. The current work addresses these limitations through the following organisation. Section \@ref(calculating-potential-cycling-demand)
focuses on calculating potential cycling demand. Section \@ref(routing)
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\@ref(road-segment-prioritization) outlines a method for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
'egalitarianism' to one grounded in 'utilitarianism'.

@@

*OD* data can be obtained from a range of sources,
including GPS data, household travels surveys and census data on work
locations. In areas where observed OD data is unavailable, modelling
techniques such as spatial interaction models can be used to estimate
travel volumes between zones [@black1995spatial; @martinez2013new;
@wilson1971family]. In this paper we use open access data from the UK
census [@ONS2011flowdata], which contains aggregate statistics on
number of commuters between administrative zones --- Middle layer Super
Output Areas (MSOA) --- by mode of travel. MSOAs have an average
population of just over 8000 [@ofn2018population].^[ See
<https://wicid.ukdataservice.ac.uk/> for open access to the OD data.]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling)
illustrate the proportion of trips cycled by distance and the
geographic extent of the input OD dataset used in this paper.

@@

\noindent where $d$ and $s$ are the distance and slope associated with each
OD pair, and $\alpha$ and all $\beta$s are parameters calculated by a regression
model on the training data. The square and square-root distance terms
"capture the non-linear impact of distance on the likelihood of
cycling", and interaction terms capture the combined effect of slope
and distance [@lovelace2017propensity]. Alternative cycling uptake
models could be 'plugged in' to our approach for different contexts or scenarios of change.

@@

For the purposes of this paper we created a similar weighting profile
that is adjusted to favor less stressful roads (based on the definition
by @sorton1994bicycle and information from Table
\@ref(table:osmroadtypes)) and roads with existing cycling
infrastructure. We believe this to be more appropriate than the simplistic
approach adopted by @mauttone2017bicycle as it makes use of a hierarchy
of road preference based on perceived stress levels, going beyond simply
favouring roads with existing cycling infrastructure.

@@

A range of algorithms could be used for prioritization. Because policy
priorities vary, we present two algorithms. Both utilize existing
infrastructure from the beginning and allow us to compare a solution
that focuses on utilitarianism to one that focuses on egalitarianism. In
both algorithms, links are selected iteratively and the iteration at
which each link is added to the solution is recorded. Investments in
cycling infrastructure can be limited by budget constraints, so it can
be useful to see where best to allocate a defined length of segregated
infrastructure. In order to incorporate egalitarian principles in our
approach, we use community detection to partition the study area and
distribute investment over the different subdivisions.

@@

Consistent growth can also be seen for the size of the Largest Connected
Component in the proposed bicycle network (Figure
\@ref(fig:componentsandGCC)). Here, however, we find that there is little
difference between both algorithms.
